An Efficient Minibatch Acceptance Test for MCMC

### 2016-07-05 nips_2016.tex
updates to first section

Markov chain Monte Carlo (MCMC) sampling is a powerful method for
computation on intractable distributions. MCMC algorithms propose
samples from a proposal distribution $q$ which is in general different
from the target distribution $p$, and decide whether to accept or
reject them based on an acceptance test. The acceptance test is
usually a Metropolis
test~\cite{Metropolis1953,hastings70}. For example, for Bayesian
posterior inference, the posterior distribution $p(\theta \mid x_1,
\ldots, x_N) \propto p(\theta)\prod_{i=1}^N p(x_i \mid \theta)$ of
parameter $\theta$ based on conditionally independent data
$\{x_i\}_{i=1}^N$ is the target distribution which one aims to
approximate by sampling $\theta$ values. When $N$ is large, it is
very expensive to compute the exact posterior because the
likelihood must be evaluated on every $x_i$.

Many state-of-the-art machine learning methods, and deep learning in
particular, are based on minibatch updates (such as SGD) to a model.
Minibatch updates produce many improvements to the model for each pass
over the dataset, and therefore have high sample efficiency. They also
map very well onto hardware such as GPUs. In contrast, MCMC methods
typically require calculations over the full dataset to produce a new
sample.  Recent results
from~\cite{cutting_mh_2014,icml2014c1_bardenet14} perform approximate
(bounded error) acceptance tests based on the classical M-H test. The
tests depend on minibatch statistics, and on the value of an
additional random variable. The amount of data consumed for each test
varies significantly from one minibatch to the next, and
non-repeatably on the random variable. By contrast,
\cite{conf/uai/MaclaurinA14} performs exact tests but requires a lower
bound on parameter distribution across its domain. The amount of data
reduction depends on the accuracy of this bound, and such bounds are
only available for relatively simple distributions.

Here we derive a new test which incorporates the variability in
minibatch statistics as {\em a natural part of the test}. Because of
this, the amount of data required for each test is fixed with high
probability and in expected value. Our test requires a bound on the
(relative) variance of the log acceptance probability ratio. This variance
changes slowly on typical problems, and we can estimate it before
performing the test.

If the variance test fails, there are three
ways to mitigate the situation:
\begin{itemize}

@@

\item Increase the minibatch size. Relative variance scales as $1/k$ vs the
  minibatch size $k$.

\item For continuous probability distributions, reduce the proposal step
  size. Since the condition is on variance in the {\em difference} in
  log probabilities of prior and proposed likelihoods, smaller proposal steps
  will lead to smaller changes in probability and smaller changes in the
  variance of those changes.
\end{itemize}

Note that changes to the distribution temperature may only need to be
temporary.  The distribution can be annealed from high-to-low
temperatures to maintain the variance condition. High temperature is
natural with small minibatches and large step sizes, since the
precision of the acceptance test is limited by the amount of data
available. As the distribution reaches a stationary state and is cooled,
particle motion can be slowed with larger minibatches or smaller proposal
steps. The contributions of this paper are as follows:

\begin{itemize}
\item We developed a new minibatch acceptance test which satisfies detailed balance.
\item we developed effectively measurable bounds for both absolute and relative
  test error. 
\item We compared performance of our new test and prior approaches. 
\item We did experiments using the approach for posterior estimation.
\end{itemize}

@@

In standard MCMC methods~\cite{gilks1996markov,brooks2011handbook}, a difficult-to-compute
probability distribution $p(\theta)$ is sampled using a Markov chain $\theta_1,\ldots,\theta_n$. The 
sample $\theta_{t+1}$ at time $t+1$ is generated using a candidate $\theta'$ from a (simpler) proposal
distribution $q(\theta'\mid \theta_t)$, filtered by an acceptance test. The acceptance test is usually
a Metropolis test. The Metropolis test has acceptance probability:
\begin{equation}\label{eq:traditional}
\alpha(\theta',\theta_t) = \min\left\{ 1, \frac{p(\theta')q(\theta_t \mid \theta')}{p(\theta_t)q(\theta' \mid \theta_t)}
\right\}
\end{equation}
and with probability $\alpha(\theta',\theta_t)$, we accept $\theta'$
and set $\theta_{t+1} = \theta'$, otherwise set
$\theta_{t+1}=\theta_t$.  The test is often implemented with an
auxiliary random variable $u \sim \mathcal{U}(0,1)$ with a comparison $u < \alpha(\theta',\theta_t)$

The acceptance test guarantees detailed balance, so that
\begin{equation}
  p(\theta)q(\theta',\theta)\alpha(\theta',\theta) = p(\theta')q(\theta,\theta')\alpha(\theta,\theta')
\end{equation}
The detailed balance condition, together with ergodicity, guarantee that the
Markov chain has a unique stationary distribution $\pi(\theta) = p(\theta)$.

For Bayesian inference, we would like to sample from a parameter distribution for $\theta$ based on some
observed data $x_1,\ldots,x_N$. i.e. we want to sample from 
$p(\theta \mid x_1, \ldots, x_N)$. The acceptance probability is now:
\begin{equation}
\alpha(\theta',\theta_t) = 
\min\left\{ 1, \frac{p(\theta')\prod_{i=1}^N p(x_i \mid \theta')q(\theta_t \mid
\theta')}{p(\theta_t)\prod_{i=1}^N p(x_i \mid \theta_t)q(\theta' \mid\theta_t)} \right\},
\end{equation}
where $p(\theta)$ is a prior, and $p(x_i \mid \theta)$ are the
probabilities of the observations. Computing samples this way requires
the use of all $N$ training data points. Moreover, it is difficult to
design tests using substantially fewer than $N$ points that also
satisfy detailed balance. To reconcile these competing
objectives,~\cite{cutting_mh_2014,icml2014c1_bardenet14} perform
approximate Metropolis-Hasting tests using sequential hypothesis
testing. During each iteration, they start with a small minibatch of
data and test the hypothesis that the sample $\theta'$ based on an
approximate version of the test $u < \alpha(\theta',\theta_t)$. If the
approximate test cannot make a decision with sufficient confidence,
then the minibatch size is increased and the test repeats. This
process continues until a decision. The bounds depend on either a
asymptotic central limit theorem \cite{cutting_mh_2014} or a
concentration bound \cite{icml2014c1_bardenet14}, the latter requiring
direct bounds on the log likelihood ratio.  The problem with both
methods is the expense of resolving small differences.  In the worst
case, all $N$ data points may be needed, and as we show next this
worst case can occur about $\sqrt{N}$ times during the performance of
$N$ tests.

Following ~\cite{icml2014c1_bardenet14}, we write the test $u < \alpha(\theta',\theta_t)$ in
the equivalent form $\Lambda(\theta,\theta') > \psi(u,\theta,\theta')$ where
\begin{equation}
\Lambda(\theta,\theta') = \frac{1}{N}\sum_{i=1}^N \log\left(\frac{p(x_i|\theta')}{p(x_i|\theta)}\right)  
\hspace{0.2in} {\rm and} \hspace{0.2in}
\psi(u,\theta,\theta') = \frac{1}{N}\log\left(u\frac{q(\theta'|\theta)p(\theta)}{q(\theta|\theta')p(\theta')}\right)
\end{equation}

By the central limit theorem, we expect $\Lambda(\theta_t,\theta')$ to be approximately
Normal. To reduce computational effort, an unbiased estimate of $\Lambda(\theta,\theta')$
based on a minibatch can be used:

\begin{equation}
\Lambda^*(\theta,\theta') = \frac{1}{b}\sum_{i=1}^b \log\left(\frac{p(x_i|\theta')}{p(x_i|\theta)}\right)  
\end{equation}

The acceptance test then becomes a statistical test of the hypothesis
that $\Lambda(\theta,\theta') > \psi(u,\theta,\theta')$ by
establishing that $\Lambda^*(\theta,\theta')$ is substantially larger
than $\psi(u,\theta,\theta')$.  In ~\cite{cutting_mh_2014} an
asymptotic central limit argument was used to derive this gap, while
in \cite{icml2014c1_bardenet14} a concentration bound was used. In
both cases, the resulting tests were shown to give useful reductions
in number of samples required over using the full dataset. But no
worst-case bounds were given on average number of samples needed.

We next show that for some simple distributions, the lower
bound of average number of data instance consumed for one iteration of
~\cite{cutting_mh_2014} and ~\cite{icml2014c1_bardenet14} is
$\Omega(\sqrt{N})$, where $N$ is the number of data points.

Consider a simple model where samples $x_i$ for $i=1,\ldots,N$ are
drawn from a Normal distribution $\mathcal{N}(\theta,1)$ where
$\theta=0.5$, and with a uniform prior on $\theta$. The log likelihood ratio is
\begin{equation}
\Lambda^*(\theta,\theta') =  \frac{1}{b}\sum_{i=1}^b \log\frac{p(x_i|\theta')}{p(x_i|\theta_t)}=
  %%\frac{1}{2}\left( (x_i-\theta')^2 - (x_i - \theta_t)^2\right) =
  (\theta'-\theta_t)\left(\frac{1}{b}\sum_{i=1}^b x_i-\theta_t-\frac{\theta'-\theta_t}{2}\right)
\end{equation}
which is normally distributed over selection of the Normal samples $x_i$.
Since the $x_i$ have unit variance, their mean has variance $1/b$, and the variance $\sigma^2(\Lambda^*)$
of $\Lambda^*(\theta,\theta')$ is $(\theta'-\theta_t)^2/b$. In order to pass
a hypothesis test that $\Lambda > \psi$, we will need to establish a
large enough gap (several $\sigma(\Lambda^*)$) between $\Lambda^*(\theta',\theta_t)$
and $\psi(u,\theta',\theta_t)$.

We would like to sample from the posterior distribution of $\theta$
which is a normal distribution centered on the sample mean, and with
variance $1/N$.  The target distribution is therefore
$\mathcal{N}(\mu,1/N)$ where $\mu$ is the sample mean.  An efficient
proposal distribution has the same variance as the target distribution
$\mathcal{N}(\mu,1/N)$, so we take $q(\theta'\mid\theta_t) =
\phi((\theta'-\theta_t)N)$, where $\phi(x)$ is a Normal density
function with zero mean and unit variance. This proposal is symmetric
$q(\theta',\theta_t)=q(\theta_t,\theta')$, and we assumed uniform
prior for $\theta$, so $\psi(u,\theta',\theta_t)$ reduces to $1/N\log
u$.

\begin{lemma}\label{lem:efficienty_cutting}
  For the model above, there
  exists a fixed constant $c$ such that with probability $ \geq c$ over the joint
  distribution of $(\theta_t, \theta', u)$, the
  test from \cite{icml2014c1_bardenet14} consumes all $N$ samples. 
\end{lemma}

\begin{proof} Assume $(\theta' - \theta_t) \in \pm\frac{1}{\sqrt{N}}[0.5,1]$ and
  $\theta_t -0.5 \in \pm\frac{1}{\sqrt{N}}[0.5,1]$ with matching sign. These events
  occur with probability $p_0=2*(\Phi(1)-\Phi(0.5)) = 0.2997$ and $p_1=(\Phi(1)-\Phi(0.5))=0.14988$ respectively. 
  Choose $u_0$
  to satisfy
  \begin{equation}
    \log u_0 = -N(\theta'-\theta_t)\left(\theta_t-0.5+\frac{\theta'-\theta_t}{2}\right) 
  \end{equation}
  and rewrite the test now as
  \begin{equation}
\frac{1}{b}\sum_{i=1}^b x_i-\left(\theta_t+\frac{\theta'-\theta_t}{2} + \frac{\log u_0}{N(\theta'-\theta)}\right) > 0
  \end{equation}
  The above choice of $u_0$ ensures that the terms in parenthesis above sum to 0.5. Since the $x_i$ have mean 0.5, the
  test will never correctly succeed.

@@

 The choices above for $\theta_t$ and $\theta'$ guarantee that
  \begin{equation}
    \log u_0 \in -[0.5,1][0.75,1.5] = -[0.375,1.5]
        \end{equation}
and consider the range of $u$ values
  \begin{equation}
    \log u \in \log u_0 + [-1,0.375]
  \end{equation}
  the size of this range is at least $\exp([-2.5,-1.125]) = [0082,0.286]$ and occurs
  with probability at least $p_2=0.2044$. The samples $\theta_t$, $(\theta'-\theta_t)$
  and $u$ are drawn independently and so the total probability of these events is
  $c = p_0 p_1 p_2 = 0.009$.
\end{proof}
There is a separate line of MCMC work drawing principles from
statistical physics. By viewing random variables as particles in a
system, one can apply Hamiltonian Monte Carlo
(HMC)~\cite{mcmc_hamiltonian_2010} methods which generate high
acceptance \emph{and} distant proposals when run on full batches of
data. Recently Langevin
Dynamics~\cite{langevin_2011,conf/icml/AhnBW12} has been applied to
Bayesian estimation on minibatches of data. This simplified dynamics
uses local proposals and avoids MH tests by using small proposal steps
whose acceptance approaches 1 in the limit. However, the constraint on
proposal step size is severe, and the state space exploration reduces
to a random walk. Full minibatch HMC for minibatches was recently
described in ~\cite{sghmc_2014} which allows momentum-augmented

proposals with larger step sizes. However, step sizes are still
limited by the need to run accurately without MH tests.
% \cite{stochastic_thermostats_2014}
Our work opens the door to applying those methods with much
more aggressive step sizes without loss of accuracy. We demonstrate
this in Section~\ref{ssec:nets}.

@@

\section{A New Metropolis-Hastings Test}\label{sec:our_algorithm}

For our new MH test, we use two key values, $\Delta$ and $\Delta'$:
\begin{equation}\label{eq:deltas}
\Delta = \log \left( \frac{p(\theta')  \prod_{i=1}^N p(x_i \mid \theta')  q(\theta_t \mid
\theta')}{p(\theta_t)\prod_{i=1}^N p(x_i \mid \theta_t)  q(\theta' \mid
\theta_t)} \right);\;  \Delta' = \log \left( \frac{p(\theta') (\prod_{i=1}^n p(x_i \mid
\theta'))^{\frac{N}{n}} q(\theta_t \mid \theta')}{p(\theta_t) (\prod_{i=1}^n p(x_i \mid
\theta_t))^{\frac{N}{n}} q(\theta' \mid \theta_t)} \right)
\end{equation}
where $p$ and $q$ have similar definitions as in Equation~\ref{eq:traditional}. Thus, $\Delta$ and
$\Delta'$ are probability ratios in log space. Note that the only difference between $\Delta$ and
$\Delta'$ is that they use $N$ and $n$ data terms $x_i$, respectively, and that the latter uses an
$N/n$ scaling term. We assume $n \ll N$ (so $\Delta'$ is substantially faster to compute) and that
the $n$ data points $x_i$ represent a random minibatch without replacement.

\subsection{The Full Data Test}\label{ssec:deltas_full}

To motivate the use of $\Delta$, we turn to Lemma~\ref{lem:detailed_balance}.

\begin{lemma}\label{lem:detailed_balance}
An acceptance function $g$ such that $g(\Delta) = \exp(\Delta) g(-\Delta )$ satisfies detailed
balance. That is, $f(\theta_t)p(\theta' \mid \theta_t) = f(\theta')p(\theta_t \mid \theta')$, where
$p(\theta_y \mid \theta_x)$ is the probability of jumping from $\theta_x$ to $\theta_y$ in our
chain.
\end{lemma}

\begin{proof}
We begin by deriving $p(\theta' \mid \theta_t)$. This is equivalent to the probability of proposing
$\theta'$ and then accepting it, so $p(\theta' \mid \theta_t) = q(\theta' \mid \theta_t)g(\Delta).$
Similarly, $p(\theta_t \mid \theta') = q(\theta_t \mid \theta')g(-\Delta).$ Notice that the
probability of accepting a transition from $\theta'$ to $\theta_t$ is $g(-\Delta)$ because this
inverts the fraction inside the logarithm term of $\Delta$.  By assumption, we can expand $g(\Delta)
= \exp(\Delta)g(-\Delta)$ in $p(\theta' \mid \theta_t)$. Doing this, and combining the result with
the definition of $p(\theta_t \mid \theta')$, we get
\begin{equation}\label{eq:combined}
g(-\Delta) = \frac{p(\theta' \mid \theta_t)}{q(\theta' \mid \theta_t)\exp(\Delta)} = \frac{p(\theta_t \mid \theta')}{q(\theta_t \mid \theta')}.
\end{equation}
Rearranging terms and expanding $\exp(\Delta)$, we have
\begin{equation}\label{eq:rearrange}
\frac{p(\theta' \mid \theta_t) f(\theta_t) q(\theta' \mid \theta_t)}{q(\theta' \mid \theta_t) f(\theta') q(\theta_t \mid \theta')} = \frac{p(\theta_t \mid \theta')}{ q(\theta_t \mid \theta')}.
\end{equation}
Cancellations result in $f(\theta') p(\theta_t \mid \theta') = f(\theta_t) p(\theta' \mid
\theta_t)$. Thus, detailed balance is satisfied.
\end{proof}

As a sanity check, the standard Metropolis-Hastings acceptance function $g(\Delta) = \min\{1,
e^\Delta \} = \min\left\{1, \frac{f(\theta')q(\theta_t \mid \theta')}{f(\theta_t)q(\theta' \mid
\theta_t)}\right\}$ satisfies the condition $g(\Delta) = \exp(\Delta)g(-\Delta)$.

>>>>

Markov chain Monte Carlo (MCMC) sampling is a powerful method for
computation on intractable distributions. We are interested primarily
in large dataset applications, where the goal is to sample a posterior
distribution $p(\theta \mid x_1, \ldots, x_N)$ of parameter $\theta$,
and where the number of data instances $N$ is large.  The
Metroplis-Hastings method (M-H) generates sample candidates from a
proposal distribution $q$ which is in general different from the
target distribution $p$, and decides whether to accept or reject them
based on an acceptance test. The acceptance test is usually a
Metropolis test~\cite{Metropolis1953,hastings70}. Conventional
Metropolis-Hastings requires all $N$ data instances to generate one
posterior sample.

Many state-of-the-art machine learning methods, and deep learning in
particular, are based on minibatch updates (such as SGD) to a model.
Minibatch updates produce many improvements to the model for each pass
over the dataset, and have high sample efficiency. They also
map very well onto hardware such as GPUs. In contrast, M-H requires
calculations over the full dataset to produce a new sample.  Recent
results from~\cite{cutting_mh_2014,icml2014c1_bardenet14} perform
approximate (bounded error) acceptance tests using subsets
(minibatches) of the full dataset. The tests depend on minibatch
statistics, and on the value of an additional random variable $u$. The
amount of data consumed for each test varies significantly from one
minibatch to the next, and depends on the current sample, the proposed
sample, and on the random variable $u$. By contrast,
\cite{conf/uai/MaclaurinA14} performs exact tests but requires a lower
bound on parameter distribution across its domain. The amount of data
reduction depends on the accuracy of this bound, and such bounds are
only available for relatively simple distributions.

Here we derive a new test which incorporates the variability in
minibatch statistics as {\em a natural part of the test}. Because of
this, the amount of data required for each test is fixed with high
probability and in expected value. Our test requires a bound on the
(relative) variance of the log acceptance probability. This variance
is small when the proposal distribution is efficient, that is when samples
diffuse at an optimal rate.

The variance of the log acceptance probability can be controlled in
three ways, and these allow the test to be applied to modest-sized
minibatches. The first two allow the test to be applied when the
minibatch size is fixed a-priori.

\begin{itemize}

@@

\item For continuous probability distributions, reduce the proposal
  step size and variance (for stochastic proposals).

\item Increase the minibatch size. Relative variance scales as $1/k$ vs the
  minibatch size $k$. Increased minibatch size also reduces the error of
  the test.

\end{itemize}

The contributions of this paper are as follows:

\begin{itemize}
\item We develop a new, more efficient (in samples per test) minibatch acceptance test which incorporates
online  estimates of its error probability.
\item We analyze the test for accuracy and speed.
\item We compared performance of our new test and prior approaches on several datasets. We demonstrate the test is several orders
  of magnitude more efficient than prior work measured as data consumed per test, and that it does not suffer
  from long-tailed data requirements (up to the dataset size). 
\end{itemize}

@@

In the Metropolis-Hastings method~\cite{gilks1996markov,brooks2011handbook}, a difficult-to-compute
probability distribution $p(\theta)$ is sampled using a Markov chain $\theta_1,\ldots,\theta_n$. The 
sample $\theta_{t+1}$ at time $t+1$ is generated using a candidate $\theta'$ from a (simpler) proposal
distribution $q(\theta'\mid \theta_t)$, filtered by an acceptance test. The acceptance test is usually
a Metropolis test. The Metropolis test has acceptance probability:
\begin{equation}\label{eq:traditional}
\alpha(\theta_t,\theta') = \min\left\{ 1, \frac{p(\theta')q(\theta_t \mid \theta')}{p(\theta_t)q(\theta' \mid \theta_t)}
\right\}
\end{equation}
and with probability $\alpha(\theta_t,\theta')$, we accept $\theta'$
and set $\theta_{t+1} = \theta'$, otherwise set
$\theta_{t+1}=\theta_t$.  The test is often implemented with an
auxiliary random variable $u \sim \mathcal{U}(0,1)$ with a comparison $u < \alpha(\theta_t,\theta')$

The acceptance test guarantees detailed balance, so that
\begin{equation}
  p(\theta)q(\theta'\mid\theta)\alpha(\theta,\theta') = p(\theta')q(\theta\mid\theta')\alpha(\theta',\theta)
\end{equation}
The detailed balance condition, together with ergodicity, guarantee that the
Markov chain has a unique stationary distribution $\pi(\theta) = p(\theta)$.

For Bayesian inference, we would like to sample from a parameter distribution for $\theta$ based on some
observed data $x_1,\ldots,x_N$. i.e. we want to sample from 
$p(\theta \mid x_1, \ldots, x_N)$. The acceptance probability is now:
\begin{equation}
\alpha(\theta_t,\theta') = 
\min\left\{ 1, \frac{p(\theta')\prod_{i=1}^N p(x_i \mid \theta')q(\theta_t \mid
\theta')}{p(\theta_t)\prod_{i=1}^N p(x_i \mid \theta_t)q(\theta' \mid\theta_t)} \right\},
\end{equation}
where $p(\theta)$ is a prior, and $p(x_i \mid \theta)$ are the
probabilities of the observations. Computing samples this way requires
the use of all $N$ training data points. Moreover, it is difficult to
design tests using substantially fewer than $N$ points that also
satisfy detailed balance. To reconcile these competing
objectives,~\cite{cutting_mh_2014,icml2014c1_bardenet14} perform
approximate Metropolis-Hasting tests using sequential hypothesis
testing. During each iteration, they start with a small minibatch of
data and test the hypothesis that the sample $\theta'$ based on an
approximate version of the test $u < \alpha(\theta_t,\theta')$. If the
approximate test cannot make a decision with sufficient confidence,
then the minibatch size is increased and the test repeats. This
process continues until a decision. The bounds depend on either a
asymptotic central limit theorem \cite{cutting_mh_2014} or a
concentration bound \cite{icml2014c1_bardenet14}, the latter requiring
direct bounds on the log likelihood ratio.  The problem with both
methods is the expense of resolving small differences.  In the worst
case, all $N$ data points may be needed, and as we show next this
worst case can occur about $\Omega(N)$ times during the performance of
$N$ tests.

Following ~\cite{icml2014c1_bardenet14}, we write the test $u < \alpha(\theta_t,\theta')$ in
the equivalent form $\Lambda(\theta_t,\theta') > \psi(u,\theta_t,\theta')$ where
\begin{equation}
\Lambda(\theta_t,\theta') = \frac{1}{N}\sum_{i=1}^N \log\left(\frac{p(x_i|\theta')}{p(x_i|\theta_t)}\right)  
\hspace{0.2in} {\rm and} \hspace{0.2in}
\psi(u,\theta_t,\theta') = \frac{1}{N}\log\left(u\frac{q(\theta'|\theta_t)p(\theta_t)}{q(\theta_t|\theta')p(\theta')}\right)
\end{equation}

By the central limit theorem, we expect $\Lambda(\theta_t,\theta')$ to be approximately
Normal. To reduce computational effort, an unbiased estimate of $\Lambda(\theta_t,\theta')$
based on a minibatch can be used:

\begin{equation}
\Lambda^*(\theta_t,\theta') = \frac{1}{b}\sum_{i=1}^b \log\left(\frac{p(x_i|\theta')}{p(x_i|\theta_t)}\right)  
\end{equation}

The acceptance test then becomes a statistical test of the hypothesis
that $\Lambda(\theta_t,\theta') > \psi(u,\theta_t,\theta')$ by
establishing that $\Lambda^*(\theta_t,\theta')$ is substantially larger
than $\psi(u,\theta_t,\theta')$.  In ~\cite{cutting_mh_2014} an
asymptotic central limit argument was used to derive this gap, while
in \cite{icml2014c1_bardenet14} a concentration bound was used. In
both cases, the resulting tests were shown to give useful reductions
in number of samples required over using the full dataset. But no
worst-case bounds were given on average number of samples needed.

We next show that for some simple distributions, the lower
bound of average number of data instance consumed for one iteration of
~\cite{cutting_mh_2014} and ~\cite{icml2014c1_bardenet14} is
$\Omega(N)$, where $N$ is the number of data points.

Consider a simple model where samples $x_i$ for $i=1,\ldots,N$ are
drawn from a Normal distribution $\mathcal{N}(\theta,1)$ where
$\theta=0.5$, and with a uniform prior on $\theta$. The log likelihood ratio is
\begin{equation}
\Lambda^*(\theta_t,\theta') =  \frac{1}{b}\sum_{i=1}^b \log\frac{p(x_i|\theta')}{p(x_i|\theta_t)}=
  %%\frac{1}{2}\left( (x_i-\theta')^2 - (x_i - \theta_t)^2\right) =
  (\theta'-\theta_t)\left(\frac{1}{b}\sum_{i=1}^b x_i-\theta_t-\frac{\theta'-\theta_t}{2}\right)
\end{equation}
which is normally distributed over selection of the Normal samples $x_i$.
Since the $x_i$ have unit variance, their mean has variance $1/b$, and the variance $\sigma^2(\Lambda^*)$
of $\Lambda^*(\theta,\theta')$ is $(\theta'-\theta_t)^2/b$. In order to pass
a hypothesis test that $\Lambda > \psi$, we will need to establish a
large enough gap (several $\sigma(\Lambda^*)$) between $\Lambda^*(\theta',\theta_t)$
and $\psi(u,\theta',\theta_t)$.

We would like to sample from the posterior distribution of $\theta$
which is a normal distribution centered on the sample mean, and with
variance $1/N$.  The target distribution is therefore
$\mathcal{N}(\mu,1/N)$ where $\mu$ is the sample mean.  An efficient
proposal distribution has the same variance as the target distribution
$\mathcal{N}(\theta'-\theta_t,1/N)$, so we take $q(\theta'\mid\theta_t) =
\phi((\theta'-\theta_t)N)$, where $\phi(x)$ is the Normal density
function with zero mean and unit variance. This proposal is symmetric
$q(\theta',\theta_t)=q(\theta_t,\theta')$, and we assumed uniform
prior for $\theta$, so $\psi(u,\theta',\theta_t)$ reduces to $1/N\log
u$.

\begin{lemma}\label{lem:efficienty_cutting}
  For the model above, there
  exists a fixed (independent of $N$) constant $c$ such that with probability $ \geq c$ over the joint
  distribution of $(\theta_t, \theta', u)$, the
  test from \cite{icml2014c1_bardenet14} consumes all $N$ samples. 
\end{lemma}

\begin{proof} Assume $(\theta' - \theta_t) \in \pm\frac{1}{\sqrt{N}}[0.5,1]$ and
  $\theta_t -0.5 \in \pm\frac{1}{\sqrt{N}}[0.5,1]$ with matching sign. These events
  occur with probability $p_0=2*(\Phi(1)-\Phi(0.5)) = 0.2997$ and $p_1=(\Phi(1)-\Phi(0.5))=0.14988$ respectively,
  and guarantee that $\Lambda^*(\theta',\theta_t)$ is negative, which guarantees that we can find
  a $u$ so that $\psi(u,\theta',\theta_t)$ equals the expected value of $\Lambda^*(\theta',\theta_t)$.
  Specifically, choose $u_0$
  to satisfy
  \begin{equation}
    \log u_0 = N(\Lambda^*(\theta',\theta_t)-\psi(1,\theta',\theta_t))
  \end{equation}
  which evaluates to 
  \begin{equation}
    \log u_0 = -N(\theta'-\theta_t)\left(\theta_t-0.5+\frac{\theta'-\theta_t}{2}\right) 
  \end{equation}
  and rewrite the test now as
  \begin{equation}
    \frac{1}{b}\sum_{i=1}^b x_i-\left(\theta_t+\frac{\theta'-\theta_t}{2} + \frac{\log u_0}{N(\theta'-\theta)}\right)
    \hspace{0.1in} \not\approx \hspace{0.1in} 0
  \end{equation}
  where $\not\approx$ means ``significantly different from'' under the distribution over samples of $x_i$. 
  The above choice of $u_0$ ensures that the terms in parenthesis above sum to 0.5. Since the $x_i$ have mean 0.5, the
  test will never correctly succeed.

@@

 The choices above for $\theta_t$ and $\theta'$ guarantee that
  \begin{equation}
    \log u_0 \in -[0.5,1][0.75,1.5] = -[0.375,1.5]
        \end{equation}
and consider the range of $u$ values
  \begin{equation}
    \log u \in \log u_0 + [-0.5,0.375]
  \end{equation}
  the size of the range in $u$ is at least $\exp([-2,-1.125]) = [0082,0.286]$ and occurs
  with probability at least $p_2=0.2044$. With $u$ in this range, we rewrite the test as:
  \begin{equation}
    \frac{1}{b}\sum_{i=1}^b x_i-0.5 \hspace{0.1in} \not\approx \hspace{0.1in} \frac{\log u/u_0}{N(\theta'-\theta)}
  \end{equation}
  so that the LHS has expected value zero.  Given our choice of intervals for the variables, the RHS is in the range $1/\sqrt{N}[-1,0.75]$.
  The standard deviation of the LHS given the interval constraints is at least $0.5/\sqrt{b}$. And
  so the gap between LHS and RHS is at most $2\sqrt{b/N}$ standard deviations.

  The samples $\theta_t$, $(\theta'-\theta_t)$
  and $u$ are drawn independently and so the probability of the conjunction of these events is
  $c = p_0 p_1 p_2 = 0.009$.
\end{proof}
There is a separate line of MCMC work drawing principles from
statistical physics. By viewing random variables as particles in a
system, one can apply Hamiltonian Monte Carlo
(HMC)~\cite{mcmc_hamiltonian_2010} methods which generate high
acceptance \emph{and} distant proposals when run on full batches of
data. Recently Langevin
Dynamics~\cite{langevin_2011,conf/icml/AhnBW12} has been applied to
Bayesian estimation on minibatches of data. This simplified dynamics
uses local proposals and avoids MH tests by using small proposal steps
whose acceptance approaches 1 in the limit. However, the constraint on
proposal step size is severe, and the state space exploration reduces
to a random walk. Full minibatch HMC for minibatches was recently
described in ~\cite{sghmc_2014} which allows momentum-augmented
proposals with larger step sizes. However, step sizes are still
limited by the need to run accurately without MH tests.
% \cite{stochastic_thermostats_2014}
Our work opens the door to applying those methods with much
more aggressive step sizes without loss of accuracy. We demonstrate
this in Section~\ref{ssec:nets}.

@@

\section{A New Metropolis-Hastings Acceptance Test}\label{sec:our_algorithm}

For our new M-H test, we denote the exact and approximate log likelihood ratios as $\Delta$ and $\Delta^*$.
First, $\Delta$ is defined as:
\begin{equation}\label{eq:delta0}
    \Delta(x,\theta_t,\theta') =
\log \frac{p(\theta')\prod_{i=1}^N p(x_i \mid \theta')q(\theta_t \mid
  \theta')}{p(\theta_t)\prod_{i=1}^N p(x_i \mid \theta_t)q(\theta' \mid\theta_t)}
  \end{equation}
and we can separate out terms dependent and independent of the data $x$ as:
\begin{equation}\label{eq:delta1}
\Delta(x,\theta_t,\theta') = N\psi(1,\theta_t,\theta') + \sum_{i=1}^N\log\frac{p(x_i\mid\theta')}{p(x_i\mid\theta_t)}
\end{equation}
\begin{equation}\label{eq:delta2}
    \Delta^*(x,\theta_t,\theta') =
N\psi(1,\theta_t,\theta') + \frac{N}{b}\sum_{i=1}^b\log\frac{p(x_i\mid\theta')}{p(x_i\mid\theta_t)}
\end{equation}
Note that $\Delta$ and $\Delta^*$ are evaluated on the full dataset and a minibatch of size $b$
respectively. The scaling term $N$ produces true log likelihood ratios whereas $\Lambda-\psi$
is a scaled log ratio. $\Delta^*$ is an unbiased estimator of $\Delta$.

\subsection{A Full Data Acceptance Test}\label{ssec:deltas_full}

The key to our test is to smooth the acceptance function.  We consider tests other than the Metropolis
test that satisfy the conditions needed for accurate posterior estimation. This condition is
the detailed balance condition defined earlier. Expressed in terms of $\Delta$, the
classical Metropolis test is $f(\Delta) = {\rm min}(\exp(\Delta),1)$. More generally we have:

\begin{lemma}\label{lem:detailed_balance}
An acceptance function $g$ such that $g(\Delta) = \exp(\Delta) g(-\Delta )$ satisfies detailed
balance. That is, $f(\theta_t)p(\theta' \mid \theta_t) = f(\theta')p(\theta_t \mid \theta')$, where
$p(\theta_y \mid \theta_x)$ is the probability of jumping from $\theta_x$ to $\theta_y$ in the
chain.
\end{lemma}

\begin{proof}
First note that the transition probability $p(\theta' \mid \theta_t)$ is the probability of proposing
$\theta'$ and then accepting it, so $p(\theta' \mid \theta_t) = q(\theta' \mid \theta_t)g(\Delta).$
Similarly, $p(\theta_t \mid \theta') = q(\theta_t \mid \theta')g(-\Delta).$ Notice that the
probability of accepting a transition from $\theta'$ to $\theta_t$ is $g(-\Delta)$ because this
inverts the fraction inside the logarithm term of $\Delta$.  By assumption, we can expand $g(\Delta)
= \exp(\Delta)g(-\Delta)$ in $p(\theta' \mid \theta_t)$. Doing this, and combining the result with
the definition of $p(\theta_t \mid \theta')$, we get
\begin{equation}\label{eq:combined}
g(-\Delta) = \frac{p(\theta' \mid \theta_t)}{q(\theta' \mid \theta_t)\exp(\Delta)} = \frac{p(\theta_t \mid \theta')}{q(\theta_t \mid \theta')}.
\end{equation}
Rearranging terms and expanding $\exp(\Delta)$, we have
\begin{equation}\label{eq:rearrange}
\frac{p(\theta' \mid \theta_t) f(\theta_t) q(\theta' \mid \theta_t)}{q(\theta' \mid \theta_t) f(\theta') q(\theta_t \mid \theta')} = \frac{p(\theta_t \mid \theta')}{ q(\theta_t \mid \theta')}.
\end{equation}
Cancellations result in $f(\theta') p(\theta_t \mid \theta') = f(\theta_t) p(\theta' \mid
\theta_t)$. Thus, detailed balance is satisfied.
\end{proof}

As a sanity check, the standard Metropolis-Hastings acceptance function $g(\Delta) = \min\{1,
e^\Delta \} = \min\left\{1, \frac{f(\theta')q(\theta_t \mid \theta')}{f(\theta_t)q(\theta' \mid
  \theta_t)}\right\}$ satisfies the condition $g(\Delta) = \exp(\Delta)g(-\Delta)$. In fact it is
the optimal acceptance function in terms of acceptance rate, since it accepts with probability
1 for $\Delta > 0$.
